Fitting metabolic models to dissolved oxygen data: The Estuarine BAyesian Single-station Estimation method (EBASE)

### 2024-03-04 revision/review_response.qmd

---
format: 
  docx:
    reference-doc: response_styles.docx
bibliography: ../refs.bib
csl: ../limnology-and-oceanography.csl

execute: 
  echo: false
  warning: false
  message: false
---

```{r}
library(dplyr)
library(tidyr)
library(lubridate)
library(ggplot2)
library(readr)
```

```{r}
#| eval: false
spelling::spell_check_files(here::here('revision/review-response.qmd'))
```

@@

Detiding (i.e., removing the influence of tidal cycles on time series data) the DO time series substantially improved model estimates and therefore was recommended as a step prior to applying the model. However, it was not clear how the authors detided their time series (they used Weighted Regression but it is unclear how this method removes tidal influence from the predicted time series – see lines 400-402). Please add additional information on the detiding approach and which variables were detided (i.e., was only DO detided? Wouldn’t all water-related input variables, such as water temperature and salinity, need to be detided?).

*Response: The detiding approach used herein is thoroughly described in another L&O paper cited in the main text [@Beck15]. Describing this approach in detail is beyond the scope of the paper, although we have added some text for clarification. Also, the method has only been applied to dissolved oxygen data, so we have not used it for any of the remaining parameters. Clarification for both points has been added to line 399: "The results of this analysis return a predicted and detided dissolved oxygen time series, where the detided results are simply the predictions at average tidal height [see @Beck15 for details]. This method was only applied to dissolved oxygen since it has not been evaluated with any other parameters."*

How would your results be impacted by using different observed data collection frequencies such as every 5-min, 15-min, 1 hour?

*Response: The resolution of "continuous" monitoring data has not been evaluated in this study, although our guess is that more frequent sampling may produce more robust estimates. However, this may not matter given the time scales at which metabolism changes (i.e., \> 1 day). In our experience, estimates from the Odum open-water method are nearly identical using 15-minute or hourly samples. Not sure the same would apply for EBASE given the fundamental differences between the approaches. We have added a point in the discussion on line 541 that speaks to this question: "These hypotheses may also consider frequency of continuous monitoring (e.g., 15 minute vs hourly) and how this may influence the results, as this is currently unknown for EBASE."*

@@

Line 267: Consider rephrasing – “…has a structure similar to, yet noisier than, that of light efficiency…”.

*Response: Rephrased.*

Line 270-271: Wouldn't one expect the opposite? During periods of high productivity, ingassing should occur given the relatively high(er) rates of primary production. Then during periods of low productivity (i.e., outside of the growing season) that outgassing would dominate as respiration of allochthonous material continues to occur while primarily production would be light limited (assuming the example dataset used is in a temperate environment).

*Response: Yes, this is correct and can even be seen in Figure 2, e.g., late September to early October shows P \> R with negative D (ingassing), then P \< R with positive D (outgassing). We have corrected this error in the text.*

@@

Line 276-278: As pointed to earlier in the manuscript, it is unclear what is meant by the optimization period. Why are the results in Figure 3 averaged by the optimization period? Why not show the daily estimates? I liked seeing the fit of the EBASE estimates on the synthetic time series however, I would like to see the fit of a linear regression of the synthetic (observed) estimates vs. the EBASE (predicted) estimates.

*Response: Please see the response to line 181. In particular, our revision notes that only one value each for $a$, $R$, and $b$ is returned for each optimization period. The vertical lines for each point in Figure 3 are credible intervals showing uncertainty around each individual estimate. The statement in the caption that "the results are averaged at the time step of the model optimization period" is incorrect and removed. This really only applies to the synthetic results in Figure 5 to show a more clear comparison with the recovered values from EBASE.*

@@

*Response: Yes, it is possible with caveats. Please see our response to reviewer one on line 188. Specifically, additional comparisons are needed and the default uninformed priors based on estuarine literature must be reconsidered.*

Model parameter b.

I’m less familiar with this term, and the authors could do a little bit more to explain what it is. On L143, they say that it is “fairly well known for open for open-ocean conditions”, but they never actually define the term. Looking at Eq 2, it seems to relate to “gas transfer efficiency” given that it is a coefficient on the wind-speed/viscosity term. That being said, I’m also a little confused on how Eq 3 is integrated into the model framework. Eq. 3 seems to relate more closely with streams as water depth is included as a scaler for kw. So somehow the two models (one more typical for lentic systems, the other more typical for lotic systems) are integrated together. In theory this makes sense in an estuary as both wind and flow likely affect gas transfer.

*Response: Apologies for this confusion. Eq. 3 is not used in EBASE, rather it is discussed as the gas transfer used in BASE that is inappropriate for our application, justifying use of the gas exchange model in @Wanninkhof14. Line 166 was revised to make this more clear: "Based on this difference and the lack of availability of other possible drivers of the transfer velocity in coastal systems, Equation 3 is not used herein. Instead, the well-established wind- and temperature-based parameterization for gas exchange of @Wanninkhof14 was used for EBASE (Equation 2)..."*

@@

Diving into eq 3, higher water levels usually indicate higher flows and thus higher gas transfer velocity (Kw). I’m not exactly sure how temperature plays into this equation, and if it is appropriate for an estuary. It seems to suggest warm periods within a day have higher exchange rates, due to differences in kinetics at the water surface. However warmer temperatures may also reflect increased stratification and less vertical mixing (small lakes). There may be something to clean from changes in temperature that drive changes in turbulence, but simply using warmer = more exchange seems like an oversimplification. Eq 3 also includes a term related to water depth (H). In flowing waters this equation suggests that Kw increases with water level, likely a result of high-water levels = higher velocities = higher turbulence = higher kw. In an estuary, I’m not sure this relationship truly holds as water depth is not as good of a surrogate for velocity. Highest water level may be at slack tide and thus minimal water movements. It could be that change in water depth (assuming the tidal currents are strong relative to the river inputs) would be a better term to reflect instantaneous discharge. Or possibly flow could be included as a possible input parameter. But I realize this term is not typically available.

*Response: We have noted an error in our description of Equation 3 that includes the depth-dependency.  As noted by Reviewer 3, @Grace15 do not include a depth-dependency in this equation. We have since removed the dependency and revised the text accordingly, emphasizing the importance of wind for the transfer velocity in open-water systems.  Regarding temperature, we feel justified in its inclusion for gas exchange given the previously noted relationship with the Schmidt number, a known factor to be important in open-water systems.*

@@

L128: mean depth?

*Response: The software can handle a single value (e.g., mean depth) or a vector of tidal height depths equal in length to the input time series. This is expressed in the documentation: https://fawda123.github.io/EBASE/reference/ebase.html. This is also expressed in the new table showing model inputs/outputs as requested above.*

L133: Thanks for providing the units. I am curious about area and time. I’m assuming that area is surface water area and time is the interval of the dissolved oxygen data. If so, perhaps expand on these so the author knows what the temporal and horizontal spatial scales are.

*Response: Please see the new table for inputs/outputs. Although equation 1 can apply to different units of space and time, we have added those used herein in parentheses: "The equation has dimensions of moles per unit area and time (e.g., $\text{mmol}~\text{m}^{-2}~\text{d}^{-1}$)."*

@@

L143: what is b? and for what time interval is b constant? This ends up being an output (Figure 3) and it is confusing to say that it is constant. So maybe it is constant within a day? Is this a standard parameter for gas exchange? As with the other parameters, providing some type of ‘name’ would be helpful, especially if provided in a table.

*Response: Please see our response to the previous comment and the new input/output table. It's an ambiguous parameter, but really just an indication of how sensitive the gas transfer is to wind speed. The sentence was revised as "The $b$ parameter is fairly well known for open-ocean conditions primarily as an expression of the sensitivity of gas exchange to wind..."*

L164: The equation implies that Kw would increase as water level (H) increases. It seems the authors have this statement inversed or talking about measured reaeration (K). I believe this relates to streams where increasing water levels relate to higher water velocities, and thus higher turbulence and higher gas transfer velocity (kw). See General Comments.

*Response: The depth-dependency for kw described in [@Grace15] was incorrect and an error on our part.  We have removed $H$ (now $Z$) from the equation and updated the text accordingly, emphasizing the wind-dependence on the transfer velocity as a key factor influencing our choice to use the method from @Wanninkhof14: "However, open-water systems like estuaries are more susceptible to wind-based influences on gas exchange, particularly for $k_w$ [@Wanninkhof14; @vandam2019]. Based on this difference and the lack of availability of other possible drivers of the transfer velocity in coastal systems, Equation 3 is not used herein. Instead, the well-established wind- and temperature-based parameterization for gas exchange of @Wanninkhof14 was used for EBASE..."*

@@

L189: Thanks for putting together the web page. I’m curious if the data in this manuscript are included within the package itself, so it is possible to regenerate the results entirely? I’m also wondering how stable this website is and if any online information that is described in the manuscript will change? I’m not that familiar with these online reference materials, but I’m thinking of a static version (tag) that will stay with the paper so a reader 10 years from now could access the exact same materials that go along with this paper.

*Response: It's the exception and not the norm that published software remains the same after a descriptive manuscript is published. For this reason, we have added a version of EBASE that is appropriate for this paper and included the version number in the text. EBASE is also now on CRAN, the Centralized R Archive Network, where a majority of publicly available R packages are hosted, in addition to the GitHub repository. We have also added text on line 193 indicating where source materials for this manuscript (i.e., to reproduce results) can be found: "All results provided herein were generated using the EBASE R package (specific materials for this manuscript can be found at <https://github.com/fawda123/ebase-manu>)."* EDIT

@@

L210: Thank you for providing the units for these parameters. I am wondering if that authors considered using negative exponents to ensure the division symbols are properly addressed. For example, 300 mmol/m2/d. While I understand this to be mmol m-2 d-1, days could be mistakenly assumed to be in the numerator of the equation. While unlikely in this situation, the subsequent equations (L218) get even more complicated “(mmol/m2/d)/(W/m2)”, so it could be worth ensuring all the units are clear. Further, this unit has m2 twice, so they could be removed from the final unit (mmol d-1 W-1).

*Response: Point well taken, we have not been consistent in the exponent notation as the text uses positive with fractions and the figures use negative with no fractions. We have corrected all instances in the text to use negative exponents. However, we have retained the units for $a$ since this makes the relationship between $P$, $a$, and $PAR$ more clear, i.e., the units for PAR ($W~m^{-2}$) when multiplied by the units for $a$ ($\left(\text{mmol}~\text{m}^{-2}~\text{d}^{-1}\right)/\left(W~m^{-2}\right)$) produce the units for P ($\text{mmol}~\text{m}^{-2}~\text{d}^{-1}$).*

@@

L237: The upper limit of 0.502 for b. This makes sense to constrain this parameter. I also noticed that many of the estimates of b are close to this value rather than near the synthetic value for the example dataset (Figure 3). I’m wondering if anything within the model could be improved to handle this issue? See general comments.

*Response: Yes, we noticed this behavior as well.  To us, this is a reflection of the instability in the parameter estimation.  Earlier iterations of model development did not have this constraint and the $b$ parameter was often estimated at unreasonably high values, often causing the other parameters ($a$ and $R$) to also be unreasonable. We previously described some rationale for why this parameters was so poorly constrained and offer some potential solutions for future research (i.e., lines 482-490, 502-509.*

@@

L272: I’m curious about the model estimates for kw. The authors show the results for flux (D), but this will vary with concentrations and k. Wondering if there is anything else to glean from those modelled results?

*Response: Perhaps, although our focus was on the primary metabolic parameters returned by EBASE given the goal of assessing parameter recovery (of which kw is included). For sake of interest, below is a plot of D and kw for the synthetic data (data available here: <https://raw.githubusercontent.com/fawda123/BASEmetab_script/master/data/apafwoxy2.csv>).  Note that kw is directly proportional to wind, whereas D is mostly a function of kw and the difference between C and Csat.*

```{r}
#| fig-width: 8
#| fig-height: 4

fwdat <- read_csv(file = url('https://raw.githubusercontent.com/fawda123/BASEmetab_script/master/data/apafwoxy2.csv'))

fwdatcmp <- fwdat %>% 
  mutate(
    DateTimeStamp = dmy_hms(datet, tz = 'America/Jamaica'),
    Date = as.Date(DateTimeStamp),
    D = -1 * `gasex,mmol/m2/d`,
    kw = `kw,m/s`
  ) %>% 
  select(Date, D, kw) %>% 
  pivot_longer(cols = c(D, kw), names_to = 'var', values_to = 'val') %>% 
  summarise(
    val = mean(val, na.rm = TRUE),
    .by = c('Date', 'var')
  ) %>% 
  mutate(
    var = factor(var, levels = c('D', 'kw'), labels = c('D (mmol m-2 d-1)', 'kw (m s-1)'))
  )

ggplot(fwdatcmp, aes(x = Date, y = val, color = var)) +
  geom_line() + 
  geom_point() + 
  theme_minimal() + 
  facet_wrap(~var, scales = 'free_y', ncol = 1, strip.position = 'left') +
  theme(
    strip.placement = 'outside', 
    legend.position = 'none'
  ) +
  labs(
    x = NULL,
    y = NULL
  )
```

@@

L301: The units for R are listed as (mmol/m2/d)/(W/m2). Perhaps the denominator is a typo?

*Response: Yes, this was a typo, corrected.*

L541: Wondering if it would be helpful to mention somewhere in this paragraph, “it depends on the question/objective of the project”? For example, if we are interested in annual summations and comparisons among systems, it may be fine to use a longer optimization period. But if we are interested in daily rates, timing of events in relation to storms, floods, etc. we may need to use a shorter optimization period. So the researcher’s question should be considered.

*Response: Good suggestion, we have added this text to line 541 (also note the last sentence that was added in response to a general comment from reviewer 1 regarding how frequency of continuous sampling may affect the results): "These hypotheses may relate to the length of time over which potential factors affect metabolism, e.g., storm events may affect metabolism at short time scales and a shorter optimization period could be used, whereas longer optimization periods may be more desirable if interested in long-term, climate-related effects on metabolism. These hypotheses may also consider frequency of continuous monitoring (e.g., 15 minute vs hourly) and how this may influence the results, as this is currently unknown for EBASE."*

@@

L14: The first use of “unphysical” in the abstract was a little unclear without further detail. Consider a short phrase like “may lead to estimates that are physically impossible” be used to define this first.

*Response: Text was modified as suggested.*

L122: The sentence “We also attempt to constrain the gas transfer velocity using this approach but are unable, at least with the data sets used in this study.“ was unclear. Consider replacing “this approach” with a more precise statement (I assume “this approach” is the Bayesian framework, rather than a reference to the previous sentence on the test with the detided inputs?). I’m also not sure if “constrain” is the right word, as the issue was accurately estimating this parameter.

*Response: The sentence was modified to clarify the meaning. We have retained "constrain" given that it's used throughout the paper, but we provide an explicit meaning in parentheses: "We also attempt to constrain (i.e., accurately estimate) the gas transfer velocity using EBASE but are unable...".*

@@

L159 & L514: The air-water exchange isn’t depth dependent in BASE. The inputs (and metabolic results) in base are all volumetric and reaeration coefficient (K) is in per day. Depth is not provided to the model but can be used to convert the metabolic estimates afterwards.

*Response: Thank you for pointing out this error.  We have removed this from the text and revised accordingly.  Please see our response to reviewer 2 for line 164.*

@@

*Response: Fixed.*

L269: Should it be “availability” rather than “variability”? of course, a more variable pool of organic matter could also be correct.

*Response: Changed to "availability", this seems more correct.*

L352: The “right answer for wrong reasons” is interesting. It could be clearer here that even though DO is estimated accurately, the P and R estimates will be inaccurate in this scenario.

*Response: Sentence was revised as follows: "...the potential for the model to produce the right answer (accurate $DO$) for the wrong reasons (inaccurate $P$, $R$, or $D$), since $DO$ is simply a mass balance of $P$, $R$, and $D$ (Equation 1)".*

### 2024-03-04 manu-draft.qmd
remove depth-dependency justification

$$
\frac{k_w}{Z} = K \cdot 1.024^{\left(T - \bar{T}\right)}
$$ {#eq-gasexbase}

where $T$ is the instantaneous water temperature, $\bar{T}$ is the average water temperature over each 24-hour period in the oxygen time series (both in $^{\circ}$C), and $K$ is the reaeration coefficient estimated from the data. @eq-gasexbase indicates that $k_w$ increases with temperature and depth. The temperature dependence is consistent with the known decrease of $k_w$ with the Schmidt number, but the depth dependence is difficult to justify. An expectation is that $k_w$ would increase as water depth decreases for a given current speed because turbulence generated at the sediment--water interface by currents would more easily reach the surface [@Ho16]. Based on these differences and the lack of availability of other possible drivers of the transfer velocity in coastal systems, @eq-gasexbase is not used herein. Instead, the well-established wind- and temperature-based parameterization for gas exchange of @Wanninkhof14 was used for EBASE [@eq-kw], but allowing $b$, which was estimated by @Wanninkhof14 to be 0.251 $\left(\text{cm}~\text{hr}^{-1}\right)/\left(\text{m}^2~\text{s}^{-2}\right)$, to be a free parameter that may reflect spatial and temporal variability in factors other than wind and temperature that may influence transfer velocity.

@@

As described in the materials and procedures section, EBASE is an estuarine adaptation of BASE [@Grace15], which was developed for stream applications and some assumptions of the underlying model of the latter were inappropriate for the former. The most critical change made to BASE was adopting the gas transfer velocity parameterization in @Wanninkhof14, which is primarily wind-based and more appropriate than the depth-dependent formulation in BASE. EBASE also removed the temperature dependency of respiration, consistent with applications of the Odum method to other shallow estuaries [@Beck15; @Murrell18]. Overall, these changes also reduced the required inputs and many monitoring locations nationwide include the requisite data for applying EBASE (i.e., dissolved oxygen, water temperature, salinity, PAR, and wind speed, as collected by the NERRS). Further, EBASE could be applied to other non-tidal, lotic systems, although additional comparisons with existing methods are needed and the default uninformed prior distributions should be chosen carefully given the justification herein based primarily on estuarine literature [e.g., @caffrey2004].

>>>>

$$
k_w = K \cdot 1.024^{\left(T - \bar{T}\right)}
$$ {#eq-gasexbase}

where $T$ is the instantaneous water temperature, $\bar{T}$ is the average water temperature over each 24-hour period in the oxygen time series (both in $^{\circ}$C), and $K$ is the reaeration coefficient estimated from the data. @eq-gasexbase indicates that $k_w$ increases with temperature, consistent with the known decrease of $k_w$ with the Schmidt number. However, open-water systems like estuaries are more susceptible to wind-based influences on gas exchange, particularly for $k_w$ [@Wanninkhof14; @vandam2019]. Based on this difference and the lack of availability of other possible drivers of the transfer velocity in coastal systems, @eq-gasexbase is not used herein. Instead, the well-established wind- and temperature-based parameterization for gas exchange of @Wanninkhof14 was used for EBASE [@eq-kw], but allowing $b$, which was estimated by @Wanninkhof14 to be 0.251 $\left(\text{cm}~\text{hr}^{-1}\right)/\left(\text{m}^2~\text{s}^{-2}\right)$, to be a free parameter that may reflect spatial and temporal variability in factors other than wind and temperature that may influence transfer velocity.

@@

As described in the materials and procedures section, EBASE is an estuarine adaptation of BASE [@Grace15], which was developed for stream applications and some assumptions of the underlying model of the latter were inappropriate for the former. The most critical change made to BASE was adopting the gas transfer velocity parameterization in @Wanninkhof14, which is primarily wind-based and more appropriate than the formulation in BASE. EBASE also removed the temperature dependency of respiration, consistent with applications of the Odum method to other shallow estuaries [@Beck15; @Murrell18]. Overall, these changes also reduced the required inputs and many monitoring locations nationwide include the requisite data for applying EBASE (i.e., dissolved oxygen, water temperature, salinity, PAR, and wind speed, as collected by the NERRS). Further, EBASE could be applied to other non-tidal, lotic systems, although additional comparisons with existing methods are needed and the default uninformed prior distributions should be chosen carefully given the justification herein based primarily on estuarine literature [e.g., @caffrey2004].
